Whole-body connectome of a segmented annelid larva

### 2025-07-15 Manuscript.qmd
MS: update Table1, Suppl File names, upload .odt

Based on these criteria, we classified `r n_neuro_celltypes` neurons into 202 cell types (Supplementary Table 1, Video 1, Figure 3).
Most neuronal cell types are represented by only two cells in the entire body (Figure 3D).

We also categorised the remaining `r num_remaining_cells` cells in the larva.
Of these, `r n_non_neuro_celltypes` were classified into 92 non-neuronal cell types (Supplementary Table 1, Figure 3---figure supplement 1).

@@

These included epidermal cells (`r n_epidermis` cells), various pigment cells (`r n_pigment` cells of 7 types), muscle cells (`r n_MUS` cells of 53 types, described in [@jasek2022]), locomotor ciliated cells (`r n_ciliated` cells of 6 types, described in [@veraszto2017]), glial cells (`r n_glia` cells of 3 types), gland cells (`r n_gland` cells of 6 types), various support and sheet cells, nephridia, putative migratory cells, parapodial cells producing chitin chaetae (`r n_chaeta` cells) or aciculae (`r n_aciculae` cells), and the follicle cells (`r n_follicle` cells of 5 types) ensheathing them (Supplementary Table 1; Video 1; Figure 3---figure supplement 1).

@@

The annotations are hierarchical and cell groups can contain one or more differentiated cell types (Supplementary Table 1).
The remaining `r n_non_celltype` cells are either dividing cells (`r n_dividing` cells), undifferentiated cells that putatively belong to the neuronal lineage (`r n_pnb` cells) and various developing neurons with projections but no or only very few synapses (`r n_all_neuronal_skids-n_neuro_celltypes`).
These cells were not classified into cell types.

@@

Supplementary Table 1 lists all neuronal and non-neuronal cell types, their main annotations, including morphological, cell-class, segment, ganglion or sensory organ, neurotransmitter and neuropeptide phenotype, and main synaptic partners.
We also listed all literature references for previously published neuronal and non-neuronal cells and indicated which cells are reported here for the first time (105 cell types out of 294).
Interactive 3D morphological renderings of each cell type together with their main annotations can also be explored on a webpage (https://jekelylab.github.io/Platynereis_connectome/;  permanent version in [@jekely_2024_10825371], /supplements/celltype_compendium_website).
Querying by two or more annotations allows tallying the number of cells in different category (e.g. number of sensory cells with a single penetrating cilium or the number of gland cells in a particular segment)(Figure 3---figure supplement 3).

@@

To explore sensory-motor pathways in the grouped graph, we searched for the shortest directed paths from all sensory neurons to all effector cells.
We identified four sensory neurons with direct motor output (sensory-motor neurons: eyespot-PRCR3, pygPBunp, pygCirrusPUSM, hPU2l-asymPDF) and 25 premotor sensory neurons.
The maximum number of hops from sensory neurons to an effector is five.
33 sensory neurons have no synaptic paths to effectors.
19 of these are neurosecretory cells with few or no synaptic partners [@williams2017] (Figure 4---figure supplement 5).
For a graph-based visualisation of information flow, we arranged the nodes according to their relative ratio of post- and presynapses (Figure 4---figure supplement 6).
This layout indicates that sensory-evoked activity propagates from sensory neurons through interneurons and motoneurons to effectors.

@@

Some sensory neurons directly connect to muscle-motor, ciliomotor or pigment-motor neurons (e.g. eyespotPRC_R3, MS cells, SNhook, SNblunt) (Figure 6---figure supplement1, Figure 7) or to effector cells (e.g. eyespotPRC_R3, hPU2l_asymPDF)(Figure 7, Figure 4---figure supplement 5).

@@

Due to contrast and focus problems in the main dataset we had to re-image certain layers at higher resolutions, to allow tracing of neurons.
These re-imaged series were made into independent CATMAID projects.
This included five extra projects, taken at various points throughout the main dataset.
All layers including the information on lost layers, immunogold labelling and re-imaging are listed in Supplementary Table 2.
The largest of these, Plexus_HT-4_Naomi_project\_\_372-4013, consisted of 1407 layers at resolution of 2.2 nm.
This stack mostly focused on the brain plexus and the ventral nerve cord where most neurites and synapses occur.
Other projects consisted of three jump/gap regions that required not only high resolution but also better realignment.
One set contained all the immunogold labelled layers [@shahidi2015serial] that were not included in the main aligned dataset.
These layers had very low contrast due to the immunolabelling procedure and therefore required higher resolution imaging.
All projects were first created and processed in TrakEM2 and then exported as flat jpeg images into CATMAID.

@@

# Figure supplements

@@

```{r, out.width = "100%", fig.cap="**Figure 4—figure supplement 5. Sensory neurons categorised by path length to effectors.**"}
#| echo: false
knitr::include_graphics("Figures/Figure4_fig_suppl5.png")
```

```{r, out.width = "100%", fig.cap="**Figure 4—figure supplement 6. Grouped connectivity graph. ** Cell-type level connectome graph with node horizontal positions proportional to the relative ratio of incoming (postsynaptic) and outgoing (presynpatic) sites. Cell-type names are shown under the nodes. <br> Figure 4—figure supplement 6—source data 1. The network in visNetwork format saved as an R RDS source file, saved with bz2 compression. Can be loaded with read_rds(). For an interactive version of the graph see also https://jekelylab.github.io/Platynereis_connectome/network_with_IO_layout.html."}
#| echo: false
knitr::include_graphics("Figures/Figure4_fig_suppl6.png")
```

@@

Supplementary Table 1.
List of all neuronal and non-neuronal cell types, their main annotations, including morphological, cell-class, segment, ganglion or sensory organ, neurotransmitter and neuropeptide phenotype, main synaptic partners and reference.

Supplementary Table 2.
Summary of all EM layers including the information on lost layers, immunogold labelling and re-imaging.

>>>>

Based on these criteria, we classified `r n_neuro_celltypes` neurons into 202 cell types (Supplementary File 1, Video 1, Figure 3).
Most neuronal cell types are represented by only two cells in the entire body (Figure 3D).

We also categorised the remaining `r num_remaining_cells` cells in the larva.
Of these, `r n_non_neuro_celltypes` were classified into 92 non-neuronal cell types (Supplementary File 1, Figure 3---figure supplement 1).

@@

These included epidermal cells (`r n_epidermis` cells), various pigment cells (`r n_pigment` cells of 7 types), muscle cells (`r n_MUS` cells of 53 types, described in [@jasek2022]), locomotor ciliated cells (`r n_ciliated` cells of 6 types, described in [@veraszto2017]), glial cells (`r n_glia` cells of 3 types), gland cells (`r n_gland` cells of 6 types), various support and sheet cells, nephridia, putative migratory cells, parapodial cells producing chitin chaetae (`r n_chaeta` cells) or aciculae (`r n_aciculae` cells), and the follicle cells (`r n_follicle` cells of 5 types) ensheathing them (Supplementary File 1; Video 1; Figure 3---figure supplement 1).

@@

The annotations are hierarchical and cell groups can contain one or more differentiated cell types (Supplementary File 1).
The remaining `r n_non_celltype` cells are either dividing cells (`r n_dividing` cells), undifferentiated cells that putatively belong to the neuronal lineage (`r n_pnb` cells) and various developing neurons with projections but no or only very few synapses (`r n_all_neuronal_skids-n_neuro_celltypes`).
These cells were not classified into cell types.

@@

Supplementary File 1 lists all neuronal and non-neuronal cell types, their main annotations, including morphological, cell-class, segment, ganglion or sensory organ, neurotransmitter and neuropeptide phenotype, and main synaptic partners.
We also listed all literature references for previously published neuronal and non-neuronal cells and indicated which cells are reported here for the first time (105 cell types out of 294).
Interactive 3D morphological renderings of each cell type together with their main annotations can also be explored on a webpage (https://jekelylab.github.io/Platynereis_connectome/;  permanent version in [@jekely_2024_10825371], /supplements/celltype_compendium_website).
Querying by two or more annotations allows tallying the number of cells in different category (e.g. number of sensory cells with a single penetrating cilium or the number of gland cells in a particular segment)(Figure 3---figure supplement 3).

@@

To explore sensory-motor pathways in the grouped graph, we searched for the shortest directed paths from all sensory neurons to all effector cells.
We identified four sensory neurons with direct motor output (sensory-motor neurons: eyespot-PRCR3, pygPBunp, pygCirrusPUSM, hPU2l-asymPDF) and 25 premotor sensory neurons.
The maximum number of hops from sensory neurons to an effector is five.
33 sensory neurons have no synaptic paths to effectors.
19 of these are neurosecretory cells with few or no synaptic partners [@williams2017] (Table 1).
For a graph-based visualisation of information flow, we arranged the nodes according to their relative ratio of post- and presynapses (Figure 4---figure supplement 5).
This layout indicates that sensory-evoked activity propagates from sensory neurons through interneurons and motoneurons to effectors.

@@

Some sensory neurons directly connect to muscle-motor, ciliomotor or pigment-motor neurons (e.g. eyespotPRC_R3, MS cells, SNhook, SNblunt) (Figure 6---figure supplement1, Figure 7) or to effector cells (e.g. eyespotPRC_R3, hPU2l_asymPDF)(Figure 7, Table 1).

@@

Due to contrast and focus problems in the main dataset we had to re-image certain layers at higher resolutions, to allow tracing of neurons.
These re-imaged series were made into independent CATMAID projects.
This included five extra projects, taken at various points throughout the main dataset.
All layers including the information on lost layers, immunogold labelling and re-imaging are listed in Supplementary File 2.
The largest of these, Plexus_HT-4_Naomi_project\_\_372-4013, consisted of 1407 layers at resolution of 2.2 nm.
This stack mostly focused on the brain plexus and the ventral nerve cord where most neurites and synapses occur.
Other projects consisted of three jump/gap regions that required not only high resolution but also better realignment.
One set contained all the immunogold labelled layers [@shahidi2015serial] that were not included in the main aligned dataset.
These layers had very low contrast due to the immunolabelling procedure and therefore required higher resolution imaging.
All projects were first created and processed in TrakEM2 and then exported as flat jpeg images into CATMAID.

@@

# Tables and Figure supplements

```{r, out.width = "100%", fig.cap="**Table 1. Sensory neurons categorised by path length to effectors.**"}
#| echo: false
library(tinytable)
Table1 <- readRDS("data/Table1.RDS")
Table1
```

@@

```{r, out.width = "100%", fig.cap="**Figure 4—figure supplement 5. Grouped connectivity graph. ** Cell-type level connectome graph with node horizontal positions proportional to the relative ratio of incoming (postsynaptic) and outgoing (presynpatic) sites. Cell-type names are shown under the nodes. <br> Figure 4—figure supplement 5—source data 1. The network in visNetwork format saved as an R RDS source file, saved with zip compression. Can be loaded with read_rds(). For an interactive version of the graph see also https://jekelylab.github.io/Platynereis_connectome/network_with_IO_layout.html."}
#| echo: false
knitr::include_graphics("Figures/Figure4_fig_suppl5.png")
```

@@

Supplementary File 1.
List of all neuronal and non-neuronal cell types, their main annotations, including morphological, cell-class, segment, ganglion or sensory organ, neurotransmitter and neuropeptide phenotype, main synaptic partners and reference.

Supplementary File 2.
Summary of all EM layers including the information on lost layers, immunogold labelling and re-imaging.
